---
title: "Ecuador plant richness analysis"
authors: "Hazel J. Anderson"
data input: "Ecuador_BIEN_plant_cleaned.csv"
data output: ""
date: "2022-11-30"
output: 
  html_document: 
    fig_width: 10
    fig_height: 10
    fig_caption: yes
---

# Set file paths
```{r}
data_path <- file.path('G:/My Drive/IBIO 870/Final project/data/L1')
output_path <- file.path('G:/My Drive/IBIO 870/Final project/data/L2')
```

# Load required packages
```{r, warning=FALSE}
library(sp)
library(sf)
library(dplyr)
library(ggplot2)
library(scico)
library(rnaturalearth)
library(purrr)
library(ggsn)
library(fasterize)
library(raster)
library(smoothr)
library(gstat)
```

# Load cleaned BIEN data
```{r}
Ecuador_plants <- read.csv(file.path(data_path,"Ecuador_BIEN_plant_cleaned.csv"))
```

# Calculate richness
Modified from https://github.com/bioXgeo/neotropical_plants/blob/master/archive/EcuadorPlantsStackedSpeciesMaps.Rmd & https://luisdva.github.io/rstats/richness/

## Convert data into sf object
```{r}
Ecuador_plants$species <- Ecuador_plants$Scientific_name

# species and lat long data
Ecuador_plants <- Ecuador_plants %>%
  dplyr::select(species, Longitude, Latitude) %>%
  na.omit()

# convert to sf
Ecuador_plants_sf <- st_as_sf(Ecuador_plants, coords = c("Longitude", "Latitude"), crs = 4326) %>%
  group_by(species) %>%
  summarize()
```

## Load in world map
```{r}
worldMap <- ne_countries(scale = "medium", type = "countries", returnclass = "sf")
# country subset
ECpoly <- worldMap %>% filter(sovereignt == "Ecuador")

# remove galapagos islands for simplicity
ECpoly <-st_crop(ECpoly, c(xmin=-84, xmax=-75.24961, ymin=-4.990625, ymax=1.455371))
# trim the map to appropriate study area.
limsEC <- st_buffer(ECpoly, dist = 1) %>% st_bbox()

# neighboring countries
COpoly <- worldMap %>% filter(sovereignt == "Colombia")
PEpoly <- worldMap %>% filter(sovereignt == "Peru")


divpolPlot <-
ggplot() +
geom_sf(data = COpoly, color = "white") +
geom_sf(data = PEpoly, color = "white") +
geom_sf(data = ECpoly) +
coord_sf(
xlim = c(limsEC["xmin"], limsEC["xmax"]),
ylim = c(limsEC["ymin"], limsEC["ymax"])
) +
theme(
plot.background = element_rect(fill = "#f1f2f3"),
panel.background = element_rect(fill = "#2F4051"),
panel.grid = element_blank(),
line = element_blank(),
rect = element_blank()
)
divpolPlot
```

## Plot points
```{r}
plantPointsPlot <-
  ggplot() +
  geom_sf(data = Ecuador_plants_sf, pch = 20) +
  coord_sf(
    xlim = c(limsEC["xmin"], limsEC["xmax"]),
    ylim = c(limsEC["ymin"], limsEC["ymax"])
  )
plantPointsPlot
```


## Define a grid
```{r}
ECGrid <- ECpoly %>%
st_make_grid(cellsize = 0.5) %>%
st_intersection(ECpoly) %>%
st_cast("MULTIPOLYGON") %>%
st_sf() %>%
mutate(cellid = row_number())

gridPlot <-
  ggplot() +
  geom_sf(data = ECpoly) +
  geom_sf(data = COpoly, color = "white") +
  geom_sf(data = PEpoly, color = "white") +
  geom_sf(data = ECGrid) +
  coord_sf(
    xlim = c(limsEC["xmin"], limsEC["xmax"]),
    ylim = c(limsEC["ymin"], limsEC["ymax"])
  ) +
  theme(
    plot.background = element_rect(fill = "#f1f2f3"),
    panel.background = element_rect(fill = "#2F4051"),
    panel.grid = element_blank(),
    line = element_blank(),
    rect = element_blank()
  )
gridPlot
```

## Richness 
```{r}
plant_richness_grid <- ECGrid %>%
  st_join(Ecuador_plants_sf) %>%
  mutate(overlap = ifelse(!is.na(species), 1, 0)) %>%
  group_by(cellid) %>%
  summarize(num_species = sum(overlap))

# plot
plantRichEC <-
  ggplot(plant_richness_grid) +
  geom_sf(data = COpoly, color = "white") +
  geom_sf(data = PEpoly, color = "white") +
  geom_sf(data = ECpoly, fill = "grey", size = 0.1) +
  geom_sf(aes(fill = num_species), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "Plant species richness") +
  coord_sf(
    xlim = c(limsEC["xmin"], limsEC["xmax"]),
    ylim = c(limsEC["ymin"], limsEC["ymax"])
  ) +
  theme(
    plot.background = element_rect(fill = "#f1f2f3"),
    panel.background = element_rect(fill = "#2F4051"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    line = element_blank(),
    rect = element_blank()
  ) + labs(fill = "richness")
plantRichEC
```

## Saving richness map
```{r}
ggsave(filename = "Richness plot cleaned data.png", dpi = 300, device="png")
```

## Calculate Richness with raw data
```{r}
Ecuador_plants_raw <- read.csv("G:/My Drive/IBIO 870/Final project/data/L0/Ecuador_BIEN_occurrence.csv")
```

## Convert data into sf object
```{r}

# species and lat long data
Ecuador_plants2 <- Ecuador_plants_raw %>%
  dplyr::select(scrubbed_species_binomial, longitude, latitude) %>%
  na.omit()

Ecuador_plants2$species <- Ecuador_plants2$scrubbed_species_binomial

# convert to sf
Ecuador_plants2_sf <- st_as_sf(Ecuador_plants2, coords = c("longitude", "latitude"), crs = 4326) %>%
  group_by(species) %>%
  summarize()
```

## Plot points
```{r}
plantPointsPlot <-
  ggplot() +
  geom_sf(data = Ecuador_plants2_sf, pch = 20) +
  coord_sf(
    xlim = c(limsEC["xmin"], limsEC["xmax"]),
    ylim = c(limsEC["ymin"], limsEC["ymax"])
  )
plantPointsPlot
```

## Richness 
```{r}
plant_richness_grid2 <- ECGrid %>%
  st_join(Ecuador_plants2_sf) %>%
  mutate(overlap = ifelse(!is.na(species), 1, 0)) %>%
  group_by(cellid) %>%
  summarize(num_species = sum(overlap))

# plot
plantRichEC <-
  ggplot(plant_richness_grid2) +
  geom_sf(data = COpoly, color = "white") +
  geom_sf(data = PEpoly, color = "white") +
  geom_sf(data = ECpoly, fill = "grey", size = 0.1) +
  geom_sf(aes(fill = num_species), color = NA) +
  scale_fill_scico(palette = "davos", direction = -1, end = 0.9, name = "Plant species richness") +
  coord_sf(
    xlim = c(limsEC["xmin"], limsEC["xmax"]),
    ylim = c(limsEC["ymin"], limsEC["ymax"])
  ) +
  theme(
    plot.background = element_rect(fill = "#f1f2f3"),
    panel.background = element_rect(fill = "#2F4051"),
    panel.grid = element_blank(),
    legend.position = "bottom",
    line = element_blank(),
    rect = element_blank()
  ) + labs(fill = "richness")
plantRichEC
```

## Saving richness map
```{r}
ggsave(filename = "Richness plot raw data.png", dpi = 300, device="png")
```
# Variograms & Moran's I
Modified from https://rstudio-pubs-static.s3.amazonaws.com/278910_3ebade4ad6a14f8f9ac6e05eb16b5a21.html & https://space-lab-msu.github.io/MSUGradSpatialEcology/lab4_autocorrelation.html

```{r}
plantVarCloud <- variogram(num_species~1, data = plant_richness_grid, cloud = TRUE)
plot(plantVarCloud, pch=20,cex=1.5,col="black",
     ylab=expression("Semivariance ("*gamma*")"),
     xlab="Distance (m)",main = "Species Richness Ecuador Plants - Cleaned")
plantVar <- variogram(num_species~1, data = plant_richness_grid)
plot(plantVar, pch=20,cex=1.5,col="black",
     ylab=expression("Semivariance ("*gamma*")"),
     xlab="Distance (m)",main = "Species Richness Ecuador Plants - Cleaned")


plantVar2Cloud <- variogram(num_species~1, data = plant_richness_grid2, cloud = TRUE)
plot(plantVar2Cloud, pch=20,cex=1.5,col="black",
     ylab=expression("Semivariance ("*gamma*")"),
     xlab="Distance (m)",main = "Species Richness Ecuador Plants - Raw")
plantVar2 <- variogram(num_species~1, data = plant_richness_grid2)
plot(plantVar2, pch=20,cex=1.5,col="black",
     ylab=expression("Semivariance ("*gamma*")"),
     xlab="Distance (m)",main = "Species Richness Ecuador Plants - Raw")
```

```{r}
plantDirVar <- variogram(num_species~1, data = plant_richness_grid, alpha=c(0,45,90,135))
plot(plantDirVar,main = "Species Richness Ecuador Plants - Cleaned via Direction")

plantDirVar2 <- variogram(num_species~1, data = plant_richness_grid2, alpha=c(0,45,90,135))
plot(plantDirVar2,main = "Species Richness Ecuador Plants - Raw via Direction")
```

```{r}
library(ape)
```

```{r}
hist(plant_richness_grid$num_species)
hist(plant_richness_grid2$num_species)
```

```{r}
qqnorm(plant_richness_grid$num_species)
qqline(plant_richness_grid$num_species,col="green")

qqnorm(plant_richness_grid2$num_species)
qqline(plant_richness_grid2$num_species,col="green")
```

Can't do a Moran's I test because the data is not normally distributed which is an assumption of Moran's I.

